% Compile on Overleaf: New Project -> Templates -> "ACL" (acl.sty, 2023+).
% Replace the template's main .tex with this file. Final (non-review) mode.
\documentclass[11pt]{article}
\usepackage{acl}
\usepackage[numbers]{natbib}
\usepackage{times}
\usepackage{latexsym}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}

\title{Distilling a Hybrid Frontier Pipeline into an Offline Open-Weight Judge \\
for Bengali Hallucination Detection}

\author{Team ⟨TEAM NAME⟩ \\ IUT Datathon — \emph{Alikbachan} (Bengali Hallucination Detection) Phase 2 \\
  \texttt{shoaibhossain2014@gmail.com}}

\begin{document}
\maketitle

\begin{abstract}
We describe our Phase 2 system for the \emph{Alikbachan} Bengali LLM hallucination
detection challenge: given a Bengali prompt, an optional supporting passage, and a candidate
response, predict whether the response is faithful or hallucinated~\citep{halu} (binary F1 on the
hallucinated class). Our Phase 1 system (private F1$_0$ = 0.910) was a hybrid of a
feature-stacked classifier, retrieval against live Bengali Wikipedia, and a
frontier-LLM judging layer — components that cannot run in an offline kernel. For
Phase 2 we convert that pipeline into a self-contained offline system: the frontier
judging layer is distilled into a QLoRA~\citep{qlora} fine-tune of Qwen2.5-7B-Instruct~\citep{qwen25} trained on
1,608 labels the Phase 1 system produced, ensembled with a feature stack that now
trains on 6$\times$ the labeled data Phase 1 had, under a layer of exact deterministic
verifiers. We evaluate on the 299 organizer-labeled samples, the only ground truth we
hold. The judge takes no gradient updates on these rows, but they drive checkpoint selection
and threshold calibration, and the final feature stack is fitted on them, so we report
them as a \emph{development set} rather than an untouched holdout. The shipped judge reaches
F1$_0$ = 0.711 there and the data-scaled stack 0.730, combining to 0.736 before
deterministic overrides. An aggressive first fine-tune scored only 0.621, \emph{below}
the 0.699 of the un-tuned base model, and the gap is concentrated in closed-book rows.
We document label provenance and our Phase 1 reproduction mechanism in full.
\end{abstract}

\section{Task and Setting}

Each row of the dataset is a triple: a Bengali prompt (\texttt{prompt\_bn}), a candidate
response (\texttt{response\_bn}), and optionally a supporting passage
(\texttt{context}); roughly half the rows are open-book (context given) and half are
closed-book. A response is labeled 0 (hallucinated) if it is fabricated, unsupported by
the given context, or factually incorrect; the metric is binary F1 on class 0. The
Phase 1 test set contained 2{,}516 rows (1{,}361 open-book / 1{,}155 closed-book)
spanning factual QA, idiom meanings, grammar and spelling items, multiple-choice
questions, and templated arithmetic word problems, across cultural bands from global to
Bangladesh-specific facts. Phase 2 is evaluated by the organizers running our notebook
on a new held-out fold, offline, under 9 hours on a single P100 or 2$\times$T4, with
all model weights under 50\,GB.

\section{Phase 1 System (Summary)}

Our Phase 1 pipeline layered three sources of authority. (1) A stacked classifier
(XGBoost~\citep{xgboost} + logistic regression, 50/50 blend) over grounding features: multilingual NLI
entailment (mDeBERTa-XNLI,~\citealp{xnli}), extractive-QA answer agreement (XLM-R SQuAD2,~\citealp{xlmr}), multilingual-e5 embedding
cosines~\citep{e5}, and lexical overlap, with decision thresholds tuned separately for
open-book and closed-book rows. (2) Retrieval from live Bengali Wikipedia providing
evidence for closed-book rows. (3) A frontier-LLM judging layer over all closed-book
rows, cross-checked by two open 32B judges (Qwen2.5-32B, Qwen3-32B), with targeted
manual verification of disagreements; templated arithmetic was verified exactly by a
regex-and-compute solver. This scored 0.910 on the private leaderboard. Three of these
components — the frontier judge, live retrieval, and manual verification — cannot run
in an offline kernel, which motivates the Phase 2 design.

\section{Phase 2 System}

\subsection{Design principle}

Rather than approximating each online component separately, we treat the \emph{outputs}
of the Phase 1 pipeline as a teacher signal and distill~\citep{distill} them into one open-weight
student: a judge that answers ``is this response correct?'' directly. That judge is then
ensembled with the Phase 1 feature stack, retrained on the enlarged corpus. The offline
system is
\[
\text{judge } P(\text{yes}) \;+\; \text{stack } P(\text{faithful})
\rightarrow \text{regime blend} \rightarrow \text{regime threshold}
\rightarrow \text{overrides,}
\]
with blend weights and thresholds fitted separately for the open- and closed-book
regimes, because the two components dominate in opposite regimes (\S5).

\subsection{Training corpus and label provenance}

\begin{table}[t]
\centering\small
\begin{tabular}{lrrl}
\toprule
Source & Rows & 0/1 & Provenance \\
\midrule
Official samples & 299 & 136/163 & organizer labels \\
Frontier no-ctx & 1{,}123 & 541/582 & frontier LLM$^{a}$ \\
Verified ctx & 485 & 251/234 & mixed$^{b}$ \\
\bottomrule
\end{tabular}
\caption{Label sources. $^{a}$Frontier-LLM judgments of Phase 1 test rows,
cross-checked against two open 32B judges; templated-arithmetic subset verified
exactly by our solver (135/135). $^{b}$Rule-derived labels (grammar templates,
regex-verified birth years against passages) plus manually reviewed uncertain rows.
The 299 official samples receive \textbf{no gradient updates in the judge}, but they are
used for model selection and calibration \emph{and} the final feature stack is fitted on
them, so they are a development set, not held-out data. Disclosure: a minority of the verified-ctx and frontier
labels incorporate human review of Phase 1 test rows. Phase 2 is scored on a
disjoint fold, so no human labeling of the evaluated data is involved.}
\label{tab:provenance}
\end{table}

Table~\ref{tab:provenance} lists the corpus. After a consistency filter (drop rows
where both cached 32B judges confidently contradict the stored label — zero rows
fired), the training set is 1{,}608 rows (792 hallucinated / 816 faithful), all drawn
from Phase 1 test rows pseudo-labeled by the Phase 1 system. Each row yields two
training examples via Bengali and English judge-prompt templates (identical to our
Phase 1 judge prompts), giving 3{,}216 examples.

\subsection{Distilled judge}

We fine-tune Qwen2.5-7B-Instruct with QLoRA (NF4, $r{=}16$, $\alpha{=}32$, dropout
0.1, all linear projections; lr $4{\cdot}10^{-5}$ cosine; max length 1{,}536; three
epochs scheduled with the \emph{first} selected by development-set evaluation;
loss on the answer tokens only). The model is trained to answer
the Bengali words for yes/no (\emph{hyam}/\emph{na}) or English Yes/No. At
inference we score $P(\text{yes})$ from the first-token logits, summing over the
disjoint first-token id sets of the yes and no surface forms (byte-fragment tokens
shared between the Bengali words are excluded), and average the Bengali- and
English-template probabilities. Thresholds on $P(\text{yes})$ are chosen per side
(open-/closed-book) on the 299-sample development set, taking the \emph{median of the
threshold region} within 0.005 of the maximum F1$_0$ rather than the argmax, for
robustness to fold shift.

\subsection{Deterministic layer}

Three exact, content-based overrides run after the judge (increasing authority):
(1) \textbf{sample match}: a test row whose prompt appears verbatim among the 299
labeled samples is resolved by \emph{exact} normalized-answer equality against the known
faithful/hallucinated responses. We removed both substring containment (which marked a
contradictory response containing the correct span as faithful) and the inverse rule that
treated any response differing from a known faithful answer as hallucinated, since
paraphrases and fuller sentences also differ; non-matching rows are now left to the model; (2) \textbf{arithmetic verification}: 14 regex
template families parse the digit-normalized prompt, compute the true answer, and
label by numeric match (validated 135/135 against derivable Phase 1 truth);
(3) \textbf{Phase 1 reproduction cache}: a SHA-256 content-keyed (context, prompt,
response) lookup of our Phase 1 predictions. On the Phase 1 test file this reproduces
our Phase 1 submission exactly (2{,}516/2{,}516 rows). It is all-or-nothing: it applies only when \emph{every} input row matches and the row
count equals the cache, so no partial application is possible and an incidental duplicate
on a new fold cannot receive a Phase 1 label. Below that gate it is inert and the distilled
system produces every prediction. The notebook logs its hit count.

\subsection{Engineering for the offline kernel}

The notebook writes a prior-based fallback submission before any GPU work, so a valid
output exists at every point; a runtime guard skips the English-template pass if the
projected total exceeds 7.5 h. Measured on 2$\times$T4 over the 2{,}516-row Phase 1
test file, the full pipeline runs well inside the limit, and the organizers' 5{,}000-row
held-out fold projects to roughly an hour against a 9 h budget. We found during dry runs
that the current Kaggle image (PyTorch 2.10, cu128) no longer supports the P100's
compute capability 6.0, so GPU execution on P100 fails for \emph{any} PyTorch model; our
notebook catches judge failure and degrades to its prior fallback while still applying
every deterministic layer. The notebook writes a baseline submission in its first cell,
using only \texttt{pandas} and the required input path, before any asset discovery or
import; all remaining work runs inside a single exception boundary and a final cell
asserts the output schema, so the notebook completes with a well-formed submission under
any downstream failure we tested. Robustness was tested by mutation: shuffled and non-contiguous ids,
varying row counts, all three no-context encodings (NaN, \texttt{[NULL]}, empty),
numeric responses, quoted newlines, and reordered or missing columns.

\section{Results}

\begin{table}[t]
\centering\small
\begin{tabular}{lrrr}
\toprule
System (scored on the 299) & all & ctx & no-ctx \\
\midrule
Qwen2.5-7B zero-shot judge & 0.699 & 0.737 & 0.686 \\
\quad + QLoRA, 3 epochs, lr $10^{-4}$ & 0.621 & 0.739 & 0.564 \\
\quad + QLoRA, 1 epoch, lr $4{\cdot}10^{-5}$ & 0.711 & 0.747 & 0.704 \\
\midrule
Stack, trained on 1{,}608 & 0.716 & 0.784 & 0.690 \\
Stack, trained on 1{,}907 (CV) & 0.730 & 0.832 & 0.690 \\
\midrule
Blend, single global weight & 0.733 & 0.838 & 0.690 \\
Blend, per-side weights & \textbf{0.737} & 0.832 & 0.698 \\
\midrule
Phase 1 stack (299-only CV, ref.) & 0.770 & 0.88 & 0.72 \\
Phase 1 full hybrid (private LB) & 0.910 & --- & --- \\
\bottomrule
\end{tabular}
\caption{F1$_0$ on the 299 organizer labels, measured \textbf{before} the
deterministic override layer, i.e.\ for the base ensemble only. The judge receives no
gradient updates on these rows; the
1{,}907-row stack row is 5-fold cross-validated over the 299 so no row is scored by a
model that saw it. Phase 1 rows are reference points, measured differently (CV on the
299 alone, and the private leaderboard).}
\label{tab:results}
\end{table}

Table~\ref{tab:results} isolates what survives the move offline. The distilled judge
recovers the frontier layer's behaviour on open-book rows (0.739), where the evidence
needed to decide is present in the prompt, but only 0.564 on closed-book rows, where
deciding requires knowing the fact. This is the expected failure mode of distillation:
judgments transfer, knowledge does not. Scaling the stack's training data from the 299
organizer rows to 1,907 (the Phase 1 pseudo-labels plus the 299) lifts open-book F1$_0$
to 0.832; the closed-book side is unmoved at 0.690, confirming the same bottleneck from
the other direction. Blending contributes little once the stack has the extra data
(the fitted judge weight is 0.0--0.3 depending on the stack's training set), because
the two components fail on the same rows. Per question type, the blend scores 0.79 on
dates, 0.79 on arithmetic, 0.78 on multiple choice, 0.77 on idiom meanings, 0.72 on
\emph{who} questions, and 0.69 on the residual \emph{other} category, which dominates
by count ($n=166$) and is where closed-book world knowledge matters most.

\section{Analysis and Limitations}

\paragraph{What does not distil.} Our central negative result is that the closed-book
gap is not a training-recipe problem. Two independent interventions — distilling
1{,}608 frontier judgments into the student, and giving the feature stack $6\times$ more
training data — both leave closed-book F1$_0$ at 0.69 \emph{on this development set},
while open-book F1$_0$ moves from
0.784 to 0.832. A sweep over 13 stack configurations (tree depth 3--6, 300--1000
estimators, two blend weights) reproduces closed-book F1$_0$ of exactly 0.6899 in twelve
of them, so the ceiling is not model capacity either. The Phase 1 teacher answered
closed-book items using world knowledge
obtained from live retrieval and web verification; a 7B student fine-tuned on the
\emph{outputs} of that process learns the decision style, not the facts. Where the
evidence needed to decide is in the input, distillation works; where it is in the
world, it does not.

\paragraph{Fine-tuning can destroy what it cannot replace.} Table~\ref{tab:results}
contains our most useful diagnostic, though we note up front that a 299-row
development set and a single 7B student cannot establish general claims about
distillation. Under the recipes we evaluated: the \emph{zero-shot} base model already scores
0.686 on closed-book rows. Our first fine-tune (lr $10^{-4}$, 3 epochs, training loss
0.002) drove that \emph{down} to 0.564 — distillation on judgments the student cannot
justify from the input does not merely fail to add knowledge, it overwrites calibrated
uncertainty with confident imitation. Re-training at lr $4{\cdot}10^{-5}$ with
checkpoint selection by per-epoch holdout evaluation recovers 0.704, and the epoch
trajectory is monotone downward thereafter (0.704, 0.638, 0.620), so the useful signal
is exhausted within one pass. Open-book rows show no such collapse (0.737 zero-shot to
0.747), consistent with grounded verification being genuinely learnable from these
labels while closed-book factuality is not.

\paragraph{Per-side blending.} Because the two components fail in opposite regimes —
the stack is far stronger with a passage (0.832 vs.\ 0.747), the judge slightly stronger
without one (0.698 vs.\ 0.690) — a single global blend weight averages two different
optima. Fitting the weight separately per evidence regime on the holdout selects
essentially the stack for open-book rows and the judge for closed-book rows, and is the
configuration we ship.

\paragraph{A calibration caveat we did not resolve.} The stack is fitted twice: a
calibration model on the 1{,}608 pseudo-labeled rows, whose predictions on the 299 are
honest and set the blend weights and thresholds, and a prediction model additionally
fitted on the 299 that scores the test rows. Thresholds therefore transfer across two
probability scales that need not coincide, so the reported figure is not exactly the
operating point of the shipped predictor. Cross-fitting the 299 would remove this and is
the first thing we would change with more time.

\paragraph{Other limits.} Arithmetic verification was exact on the recognised Phase 1 template families and is
not evidence about unseen templates. Its regexes are not anchored to the full prompt and
units are unverified, so a false parse would become a confident hard label; we therefore
restrict it to \emph{canonical} responses (no negation, a single numeric value, at most
one day name) and skip any row that fails to parse. This narrows but does not eliminate
the risk, and it is the component we would most want independently stress-tested. unparsed math rows fall back to the
ensemble, which has learned a strong hallucinated prior for closed-book arithmetic.
Pseudo-label noise bounds the student from above; we mitigate with cross-judge
consistency filtering and exact verification where possible, and teacher accuracy was 0.95 on the 41 closed-book rows whose ground truth happened to
be derivable, a selected subset that need not represent the teacher overall.
Finally, all thresholds are calibrated on 299 rows, so we choose flat-region medians
rather than argmax values to reduce variance under fold shift.

\paragraph{Engineering findings for offline kernels.} Several environment facts cost us
dry-run cycles and are worth recording. The current Kaggle image (PyTorch 2.10, cu128)
no longer supports the P100's compute capability 6.0, so GPU PyTorch cannot run on that
accelerator at all. \texttt{bitsandbytes} is absent from the image, ruling out 4-bit
inference without a vendored wheel. \texttt{peft}~0.19 raises on a \texttt{torchao}
version check when attaching an adapter, so we merge LoRA weights directly
($W \mathrel{+}= (\alpha/r)BA$) and drop the dependency. Locating the base model by
globbing for \texttt{model*.safetensors} silently finds the wrong network once encoder
mirrors are also attached, so we match on \texttt{config.json} architecture instead.
Finally, sorting prompts by length for padding efficiency concentrates the longest
sequences in the final batches, which OOMs a fixed batch size partway through a run; we
batch to a token budget and recursively halve any batch that still fails. Because the
notebook writes a valid submission before any GPU work and treats every model stage as
optional, each of these failures degraded rather than disqualified the run.

\section{Compliance and Reproducibility}

All inference-time models are open-weight (Qwen2.5-7B-Instruct base + our LoRA
adapter; both attached as Kaggle datasets/models), no internet access is required, and
total weights are $\approx$18\,GB (15.2\,GB base, 0.15\,GB adapter, 2.8\,GB encoders). Fine-tuning on the labeled sample set and own
models are explicitly permitted by the competition rules; we additionally disclose all
label provenance in Table~\ref{tab:provenance} and the reproduction-cache mechanism in
\S3.4. The README lists the exact environment and artifacts.

\section*{Acknowledgments}
We thank the organizers for the dataset and for the Phase 2 infrastructure.

\begin{thebibliography}{9}
\bibitem{qwen25} A.~Yang et al. Qwen2.5 Technical Report. \emph{arXiv:2412.15115}, 2024.
\bibitem{qlora} T.~Dettmers, A.~Pagnoni, A.~Holtzman, L.~Zettlemoyer. QLoRA: Efficient
Finetuning of Quantized LLMs. \emph{NeurIPS}, 2023.
\bibitem{lora} E.~Hu et al. LoRA: Low-Rank Adaptation of Large Language Models.
\emph{ICLR}, 2022.
\bibitem{distill} G.~Hinton, O.~Vinyals, J.~Dean. Distilling the Knowledge in a Neural
Network. \emph{arXiv:1503.02531}, 2015.
\bibitem{e5} L.~Wang et al. Multilingual E5 Text Embeddings. \emph{arXiv:2402.05672}, 2024.
\bibitem{xgboost} T.~Chen, C.~Guestrin. XGBoost: A Scalable Tree Boosting System.
\emph{KDD}, 2016.
\bibitem{xnli} A.~Conneau et al. XNLI: Evaluating Cross-lingual Sentence
Representations. \emph{EMNLP}, 2018.
\bibitem{xlmr} A.~Conneau et al. Unsupervised Cross-lingual Representation Learning at
Scale. \emph{ACL}, 2020.
\bibitem{halu} Y.~Zhang et al. Siren's Song in the AI Ocean: A Survey on Hallucination
in Large Language Models. \emph{arXiv:2309.01219}, 2023.
\end{thebibliography}

\end{document}
